---
title: "Project Report"
author: "Orhan Kocak, Enis Arda Iskender, Kerem Efe Akyurt"
date: "2024-01-13"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(readxl)
library(tidyverse)

  unemploymentData <- read_excel("unemployment.xls", skip = 6, col_names = c("Year","Count","Percentage"))
  unemploymentData <- unemploymentData |>
    slice(0:8)
  unemploymentData[8,1] = "2021"
  
      priceIndex <- read_excel("inflation.xls", skip = 4, col_names = c("Year","January","February","March","April","May","June","July","August","September","October","November","December"))
    priceIndex <- priceIndex |>
    slice(12:19)|>
    mutate(across(2:last_col(), as.numeric))|>
    mutate(Total = rowSums(across(January:December)), .before = January)
    
      rows_to_read_male <- c(42,81,120,159,198,237,276,315)
  rows_to_read_female <- c(43,82,121,160,199,238,277,316)
  suicideData <- read_excel("suicide.xls", skip = 4)
  suicideData_male <- suicideData |>
    slice(rows_to_read_male)|>
    select(-c(1,2))|>
    mutate(Year = c(2021,2020,2019,2018,2017,2016,2015,2014), .before = Sex)|>
    arrange(-dplyr::row_number())
  suicideData_female <- suicideData |>
    slice(rows_to_read_female)|>
    select(-c(1,2))|>
    mutate(Year = c(2021,2020,2019,2018,2017,2016,2015,2014), .before = Sex)|>
    arrange(-dplyr::row_number())
  
  inflation_data  <- read.csv("inflation_oecd.csv")
  inflation_data_cleared <- inflation_data[, !(names(inflation_data) %in% c("INDICATOR","MEASURE", "SUBJECT", "FREQUENCY", "Flag.Codes"))]
  inflation_data_filtered <- inflation_data_cleared %>%
    filter(between(TIME, 2000, 2019))
  
    unemployee_data <- read.csv("unemployee_rate.csv")
  unemployee_data_oecd <- unemployee_data %>% filter(LOCATION == "OECD")
  unemployee_data_oecd_filtered <- unemployee_data_oecd[, !(names(unemployee_data_oecd) %in% c("MEASURE", "SUBJECT", "FREQUENCY", "Flag.Codes"))]
  
  suicide_data  <- read.csv("global_suicide_rates.csv", skip = 1)
suicide_data_cleared <- suicide_data %>%
    mutate(across(starts_with("X"), ~gsub("\\[.*?\\]", "", .)))
suicide_data_filtered <- suicide_data_cleared %>%
    gather(key = "year_rate", value = "rate", starts_with("X")) %>%
    mutate(year = gsub("X\\.", "", year_rate)) %>%
    separate(year, into = c("year", "X"), sep = "\\.") %>%
    select(-X, -year_rate) %>%
    arrange(year)
suicide_data_filtered_avg <- suicide_data_filtered %>%
    group_by(year, Sex) %>%
    summarise(rate = mean(as.numeric(rate), na.rm = TRUE))
suicide_data_filtered_avg_unisex <- suicide_data_filtered_avg %>%
    filter(Sex == " Male")

  plot_suicide_vs_price <- ggplot() +
    geom_bar(data = suicideData_male, aes(x = as.numeric(Year) - 0.2, y = as.numeric(Total), fill = "Male"), 
             color = "cyan", position = "dodge", stat = "identity", width = 0.4) +
    geom_bar(data = suicideData_female, aes(x = as.numeric(Year) + 0.2, y = as.numeric(Total), fill = "Female"), 
             color = "pink", position = "dodge", stat = "identity", width = 0.4) +
    geom_line(data = priceIndex, aes(x = as.numeric(Year), y = as.numeric(Total) / 2, group = 1), 
              color = "#0000aa", linewidth = 1.2) +
    scale_y_continuous(
      name = "Suicide Rates",
      sec.axis = sec_axis(~ . * 2, name = "Consumer Price", breaks = seq(0, max(priceIndex$Total), by = 500))
    ) +
    scale_fill_manual(values = c("Male" = "cyan", "Female" = "pink")) +
    theme(
      axis.title.y = element_text(color = "black", size = 12),
      axis.title.y.right = element_text(color = "#0000aa", size = 12)
    ) +
    labs(title = "Suicide Rates Vs. Consumer Price Index") + xlab("Year")
  plot_suicide_vs_unemployment <- ggplot() +
    geom_bar(data = suicideData_male, aes(x = as.numeric(Year) - 0.2, y = as.numeric(Total), fill = "Male"),
             color = "cyan", position = "dodge", stat = "identity", width = 0.4) +
    geom_bar(data = suicideData_female, aes(x = as.numeric(Year) + 0.2, y = as.numeric(Total), fill = "Female"),
             color = "pink", position = "dodge", stat = "identity", width = 0.4) +
    geom_line(data = unemploymentData, aes(x = as.numeric(Year), y = Count),
              color = "#0000aa", linewidth = 1.2, group = 1) +
    scale_y_continuous(
      name = "Suicide Rates",
      sec.axis = sec_axis(~ ., name = "Unemployment",
                          breaks = seq(0, max(unemploymentData$Count), by = 500))
    ) +
    scale_fill_manual(values = c("Male" = "cyan", "Female" = "pink")) +
    theme(
      axis.title.y = element_text(color = "black", size = 12),
      axis.title.y.right = element_text(color = "#0000aa", size = 12)
    ) +
    labs(title = "Suicide Rates Vs. Unemployment") + xlab("Year")
  
  inflation_suicide_global <- ggplot() +
  geom_line(data=suicide_data_filtered_avg_unisex,aes(x=as.numeric(year), y = (as.numeric(rate))), color="#aa0000", linewidth=1.2) +
  geom_line(data=inflation_data_filtered, aes(x=as.numeric(TIME),y=as.numeric(Value)), color="#0000aa", linewidth=1.2) +
  scale_y_continuous(
    name = "Suicide Rates",
    sec.axis = sec_axis(~., name="Inflation")
  ) +
  theme(
    axis.title.y = element_text(color = "#aa0000", size=12),
    axis.title.y.right = element_text(color = "#0000aa", size=12)
  ) +
  labs(
    title="Inflation-Suicide Plot",
    caption="Suicide count per 100.000 people*"
  ) +
  xlab("Year")
  
  unemployment_suicide_global <- ggplot() +
  geom_line(data=suicide_data_filtered_avg_unisex,aes(x=as.numeric(year), y = (as.numeric(rate))), color="#aa0000", linewidth=1.2) +
  geom_line(data=unemployee_data_oecd_filtered, aes(x=as.numeric(TIME),y=as.numeric(Value)), color="#0000aa", linewidth=1.2) +
  scale_y_continuous(
    name = "Suicide Rates",
    sec.axis = sec_axis(~., name="Unemployment")
  ) +
  theme(
    axis.title.y = element_text(color = "#aa0000", size=12),
    axis.title.y.right = element_text(color = "#0000aa", size=12)
  ) +
  labs(
    title="Unemployment-Suicide Plot",
    caption="Suicide count per 100.000 people*"
  ) +
  xlab("Year")
```

## Research Question
Is there any relation between inflation, unemployment and suicide data over the years?

## Datasets Used 
-   \- [Consumer Price Index](https://data.tuik.gov.tr/Bulten/DownloadIstatistikselTablo?p=VNahkuJlRdjGi186A0UyGFl4/vvuLqcU9WNSamrnmn2BsuZQSCYHsScr/wPRAsfd){.uri}
-   \- [Unemployment Rates](https://data.tuik.gov.tr/Bulten/DownloadIstatistikselTablo?p=8qvM/mN5mN8/qVhH4ZBw5Oevs/TvsYlRCOC2Dg/dcM7k8bk7Eod5aaqpj9EGFPTS){.uri}
-   \- [Yearly Suicide Data](https://data.tuik.gov.tr/Bulten/DownloadIstatistikselTablo?p=9u2oCqacK3ZUqx4qFHbuTulw5EqmeCn1xvvQcrq4mI95KN20YzGH7r0a1Be7Bz1y){.uri}
-   \- [Global Suicide Data](https://apps.who.int/gho/data/node.main.MHSUICIDEASDR?lang=en){.uri}
-   \- [Global Inflation Data](https://data.oecd.org/price/inflation-cpi.htm){.uri}
-   \- [Global Unemployment Rates](https://data.oecd.org/unemp/unemployment-rate.htm){.uri}

# Data Views (Cleaned)

## Unemployment Data

```{r echo=FALSE}
head(unemploymentData)
```

## Consumer Price Index

```{r echo=FALSE}
head(priceIndex)
```

## Suicide Data

```{r echo=FALSE}
head(suicideData_male)
```

## Global Inflation Data

```{r echo=FALSE}
head(inflation_data_filtered)
```

## Global Unemployment Data

```{r echo=FALSE}
head(unemployee_data_oecd_filtered)
```

## Global Suicide Data

```{r echo=FALSE}
head(suicide_data_filtered_avg_unisex)
```

# Plots

## Turkey: Suicide - Consumer Price Index
```{r echo=FALSE}
plot_suicide_vs_price
```

Results show there might be a correlation between consumer prices and suicides, and steadily growing.

## Turkey: Suicide - Unemployment
```{r echo=FALSE}
plot_suicide_vs_unemployment
```

Results show there is no relation between unemployment and suicides.

## World: Suicide - Inflation
```{r echo=FALSE}
inflation_suicide_global
```

## World: Suicide - Unemployment
```{r echo=FALSE}
unemployment_suicide_global
```